Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-11 POPL/paper.tex
updates to intro

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.
%
The final frontier for parsing research is the development of
\emph{formally verified} parser implementations, which due to the
pervasive nature of parsing would be a useful component of many
formally verified software systems: compilers and servers most
especially.
%
\steven{Some ideas on what the unifying message throughout the paper should be:
we offer an abstraction over the formalisms used ad-hoc when doing formal
language/grammar theory. Whether a user of these constructs recognizes it or not, there is some
universality between all of their systems. We don't necessarily do much, per se, other
than abstract over the commonalities and provide a common language.

However, just because there isn't a ton of clever work in designing this syntax, as
a lot of it does flow very naturally from the literature, that doesn't diminish
the value of the syntax. In fact, it emboldens the syntax because we want to
appeal to the idea that ``there couldn't have been any other choice for
universal grammar''.

That is all to say, our appeal is generalization. I think does appear
throughout the paper, but is maybe too aimless of a message to be satisfying.
Perhaps, we should also lean on applications more. Idk I'm gonna sleep now, and
we'll talk in the morning
}

In this work we aim to refine formal language theory into a
mathematical theory of formal grammars that can be used to develop
\emph{correct-by-construction} parsers. Our approach can be summarized
in the Curry-Howard-style slogan ``grammars as types'', and the
addendum ``parsers as terms'' where since the types of the terms are
grammars, this means not just parsers but intrinsically verified
parsers as terms. That is, we develop a type theory with a novel
denotational semantics in a category with abstract formal grammars as
objects and \emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyan grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich
category: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

@@

The substructural nature of \theoryabbv is well-aligned with the
requirements intrinsic to parsing and to the theory of formal
languages, where strings constitute a very clear notion of resource
that cannot be duplicated, reordered or dropped.

%% The theory of formal languages and parsing is one of the oldest and
%% most thoroughly developed areas of theoretical computer science. A
%% prominent topic in the 1950, 60s, and 70s led to a series of remarkable
%% developments: 
In this paper, we propose a structured generalization of the
theoretical machinery used for parsing. In conjunction with an updated
point-of-view on formal grammars, we provide a domain-specific
language to build and verify parsers.

\steven{The above paragraph maybe isn't the right thing to say, but I want
  something that ties the first paragraph to the contributions that follow}

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
including one to prove a metatheoretic canonicity theorem. Finally, in
\cref{sec:automata} we give a presentation of automata as grammars. With this
view in mind, we may encode parsing algorithms and execute them entirely
internal to our formalism. Our contributions are then:
%
\begin{itemize}
  \item A syntax-independent notion of grammar given by the category $\Grammar$
  \item \theoryname: A dependent
    linear-non-linear type theory meant to syntactically capture the latent structures
    in $\Grammar$ that enable parsing
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
  \item An implementation of the language, which is embedded shallowly in Agda
\end{itemize}

@@

The notion of formal language is central to the theory of parsing. A
\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
\emph{subset} of strings $L \subseteq \String$. This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.
In this section we propose a \emph{proof-relevant} generalization of formal
languages that can accommodate richer aspects of formal language theory such as
parsing. We begin by proposing a presheaf approach to grammars that encompasses
other classes of grammars in a syntax-independent way.

\paragraph{From Formal Languages to Formal Grammars}
A recognizer should be extensionally equivalent to the indicator function
$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
that it is in the language $w \in L$. However, language recognition is
insufficient for specifying a parser --- a function that decomposes a string
into semantic components that adhere to the structure of a formal grammar. In
practice, parsers do more than simply return whether a string is valid or
invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
  trees are usually not materialized in memory, as semantic actions often perform
  computation over the trees while they are being constructed.}.

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}

@@

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
%
\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
%
%
\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.
%
\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
%
\pedro{We should justify, even if briefly, the existence of this
  fixed-point}

>>>>

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.

The theory of formal languages and parsing is one of the oldest and most
thoroughly developed area of theoretical computer science.
The central object of study is a \emph{formal language} $L$ over an alphabet
$\Sigma$, which is classically defined as a
\emph{subset} of strings $L \subseteq \String$. The expressivity of formal
languages is sufficient for defining a \emph{recognizer} of a language $L$:
a function $r : \String \to \Prop$ such that $r$ maps each string $w$ to the
proposition that $w$ is in the language $w \in L$. This technique of language
recognition can concisely specify lexical analysis, and even lead to verified
lexer implementations \cite{egolfVerbatim,Ouedraogo_2023}.

While there are ample thoery and tooling to handle the first stage of a language frontend, the final
frontier for parsing research is the development of
\emph{formally verified} parser implementations. Due to the
ubiquitous nature of parsing, these would useful components of many
formally verified software systems: compilers and servers most
especially. However, language recognition is insufficient for specifying a
parser, and for this task much of the structured foundations from the theory of
formal languages is rendered moot. Rather than returning a propositon-valued result, a parser produces a
structured \emph{parse tree} adherent to a \emph{formal grammar}.
The research on such grammars is often performed atop an ad-hoc choice of grammar
formalism, and results are proven with respect to that formalism. Critically,
the field is lacking a unifying theory across these different formalisms, even
though many of the same
compositional reason principles appear across distinct grammar formalisms.

In this work, we aim to elucidate these similarities and
provide a mathematical theory of formal grammars for constructing verified
parsers. Our approach can be summarized with the
Curry-Howard-style slogan: ``grammars as types''; and the addendum: ``parsers as
terms''. That is, we develop a type theory with a novel
denotational semantics in a category with abstract formal grammars as
objects and \emph{parse transformers} as morphisms. We say these are
\emph{abstract} formal grammars because they are not tied to any
particular syntactic formalism such as Chomskyan grammars, regular
expressions, Lambek calculus etc. The category we choose is a rich
in structure: it is a topos as well as being monoidal closed, and these
universal properties serve as the inspiration for our type theory.

@@

The substructural nature of \theoryabbv is well-aligned with the
requirements intrinsic to parsing and to the theory of formal
languages, where strings constitute a very clear notion of resource
that cannot be duplicated, reordered, or dropped. Moreover, the constructive
aspect of \theoryabbv ensures that verification of parsers written in the
calculus are \emph{correct-by-construction}. The type system is powerful enough
that derivations of a term type checking carry intrinsic proofs of correctness.

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. In \cref{sec:automata} we demonstrate the usage of
\theoryname for defining parser terms and internalizing results from formal language theory.
Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards in \cref{sec:othermodels}, we demonstrate several models,
including one to prove a metatheoretic canonicity theorem.
Our contributions are then:
%
\begin{itemize}
  \item A syntax-independent notion of grammar given by the category $\Grammar$
  \item \theoryname: A dependent
    linear-non-linear type theory meant to syntactically capture the latent structures
    in $\Grammar$ that enable parsing
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Grammar$
  \item An implementation of the language, which is embedded shallowly in Agda
\end{itemize}

@@

The notion of formal language is central to the theory of parsing.
This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.
In this section we propose a \emph{proof-relevant} generalization of formal
languages that can accommodate richer aspects of formal language theory such as
parsing. We begin by proposing a presheaf approach to grammars that encompasses
other classes of grammars in a syntax-independent way.

\paragraph{From Formal Languages to Formal Grammars}

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.

Note, to ensure that the linear function types do indeed generalize Brzozowski
derivatives, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}

@@

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
%
\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
%
%
\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.
%
\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
%
\pedro{We should justify, even if briefly, the existence of this
  fixed-point}
\steven{As a consequence of Knaster-Tarski}
